---
title: "Denver: OpenStreetMap against the city's own inventory"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Denver: OpenStreetMap against the city's own inventory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

This walk-through uses `denver_lodo`, a frozen sample of both data sources
for a small box over downtown Denver, so it runs without network access.
The last section shows the same calls at city scale.

```{r setup}
library(cyclelanes)
```

## 1. Classify the OpenStreetMap ways

`denver_lodo$osm` is exactly what `cl_fetch_osm()` returns: one row per
way with whatever OSM tags exist in the area, unclassified.

```{r}
nrow(denver_lodo$osm)
head(names(denver_lodo$osm), 20)
```

`cl_classify()` turns the tagging into the package taxonomy, resolving
`cycleway:left` / `cycleway:right` per side and keeping the more protected
side as `facility_type`. Ways that carry a bicycle tag but are not a
facility (a sidewalk with `bicycle=designated`, a road with `cycleway=no`)
are dropped with `drop_none = TRUE`.

```{r}
lanes <- cl_classify(denver_lodo$osm, drop_none = TRUE)
knitr::kable(cl_summary(lanes), digits = 2)
```

```{r, eval = has_ggplot2, fig.alt = "Map of downtown Denver bike facilities from OpenStreetMap, coloured by facility type"}
cl_plot(lanes, title = "Downtown Denver, OpenStreetMap")
```

## 2. The city's inventory through the same taxonomy

`denver_lodo$official` is the same box from Denver's *Denver Moves: Bikes*
2025 layer, already crosswalked into the taxonomy. The city's own class is
kept in `official_class`, and `status` says whether a segment is built as
recorded or has upgrades planned.

```{r}
knitr::kable(cl_summary(denver_lodo$official, by = c("facility_type", "status")), digits = 2)
```

## 3. How well do they agree?

`cl_compare()` measures two things. Spatially: what share of each layer
lies within `tolerance` metres of the other. And by type: for every
matched segment, which facility type the other layer assigns to it.

```{r}
cmp <- cl_compare(lanes, denver_lodo$official, tolerance = 15)
cmp
```

The confusion matrix is where the story is. Rows are the city's type,
columns are what OSM calls the same ground, in kilometres.

```{r}
round(cmp$confusion, 1)
```

Downtown, both sources agree almost everywhere that *something* exists,
and on *what* for roughly three quarters of the length. Most of the
remaining disagreement is buffered versus painted lanes, which is often a
tagging choice rather than a difference on the ground; `type_adjacent`
counts those as near-misses.

One convention mattered a lot here. Denver's mappers draw the downtown
protected lanes as their own `highway=cycleway` ways beside the road,
named after the street ("14th Street", "Lawrence Street"), rather than as
a `cycleway=track` tag on the road. Read literally those are off-street
trails, and before the classifier learned to recognise street-named
cycleways as sidepaths, OSM showed 7 km of protected lane downtown
against the city's 11 km and type agreement was near 50%. The rule is
`sidepath_by_name`, on by default and documented in `?cl_classify`.

```{r, eval = has_ggplot2, fig.alt = "Map of the city's downtown segments coloured green where OpenStreetMap assigns the same facility type, red where it differs, grey where OSM has nothing nearby"}
cl_plot(cmp$official, colour = "type_match",
        title = "City segments: does OSM agree on the facility type?")
```

Segments the city has and OSM lacks entirely are the ones with a low
matched fraction:

```{r}
gaps <- cmp$official[cmp$official$matched_frac < 0.5, ]
nrow(gaps)
knitr::kable(sf::st_drop_geometry(gaps)[, c("name", "official_class", "status")])
```

## 4. How stressful is it to ride?

`cl_lts()` turns facility type plus the parent road's speed and lane
count into a Level of Traffic Stress from 1 (almost anyone) to 4 (strong
and fearless). Missing tags are filled from the road class and recorded
in `lts_basis`.

```{r}
stress <- cl_lts(lanes)
knitr::kable(cl_summary(stress, by = "lts"), digits = 2)
head(table(stress$lts_basis), 6)
```

```{r, eval = has_ggplot2, fig.alt = "Map of downtown Denver bike facilities coloured by Level of Traffic Stress from green (1) to red (4)"}
cl_plot(stress, colour = "lts", title = "Downtown Denver: Level of Traffic Stress")
```

## 5. The same thing at city scale

Everything above works on a whole city; it just needs the network. A place
name is geocoded, the Overpass box is clipped to the city's administrative
boundary so Lakewood and Aurora do not leak in, and the official layer is
cut to the same polygon.

```{r, eval = FALSE}
denver <- cl_bike_lanes("Denver, Colorado", backend = "extract", cache = TRUE)
official <- cl_fetch_official("denver", bbox = attr(denver, "cl_boundary"))
cl_compare(denver, official)
```

Run on 2026-09-03 against the Colorado Geofabrik extract, that gives
5,407 OSM facilities and 642 km inside the city limits against the city's
811 segments and 616 km. Spatially the two agree on 78% of each other's
length; on facility type, 55%, or 87% once adjacent levels count. The
confusion matrix shows two conventions rather than errors: the city files
169 km of off-street network as "Trail" where OSM tags most of it as
`path` + `bicycle=designated` (`shared_use_path` here, 261 km of it), and
the city's 62 km of neighborhood bikeways barely exist in OSM (5 km),
because a quiet street with wayfinding signs has little for a mapper to
see.

Two things to keep in mind at that scale:

* **Use the extract backend for a city.** Overpass is a shared public
  server that rate-limits by address and refuses long queries when busy;
  `cl_bike_lanes()` retries with backoff and `cache = TRUE` avoids
  re-fetching, but `backend = "extract"` sidesteps it entirely by reading
  a Geofabrik extract through `osmextract`. The first download of a state
  is large (Colorado is 260 MB); after that every city in it is local.
* **`shared_use_path` is where OSM is generous.** Park footways tagged
  `bicycle=designated` count; `strict = TRUE` drops the ones with no paved
  surface tag when you want on-road-quality kilometres. Denver strict:
  3,436 facilities and 388 km.

## Provenance

OpenStreetMap data is © OpenStreetMap contributors, ODbL. The Denver layer
is the City and County of Denver's Denver Moves: Bikes 2025 bikeway
inventory, via the Denver Open Data Catalog. `denver_lodo$fetched` records
when both were fetched: `r denver_lodo$fetched`.
